\textbf{ Interpretable Gaussian Processes for Stellar Variability }

### 2020-12-05 tex/gitlinks.tex
more text; fix test stability issue

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/c50415978a292c49f67eb6752e3f57c606f8bf6f/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/c50415978a292c49f67eb6752e3f57c606f8bf6f/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/c50415978a292c49f67eb6752e3f57c606f8bf6f/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

>>>>

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/41f024ad5662f691f48103c4041eb3e915523e38/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/41f024ad5662f691f48103c4041eb3e915523e38/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/41f024ad5662f691f48103c4041eb3e915523e38/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

### 2020-12-05 tex/ms.tex
more text; fix test stability issue

Over the past decade, Gaussian processes (GPs) have gained traction as
a leading tool for modeling correlated signals in astrophysical timeseries.
Despite whatever mystique the words ``Gaussian process'' may evoke, a GP
is nothing but a Gaussian distribution in many (formally infinite)
dimensions. Specifically, it is a Gaussian distribution over
\emph{functions} spanning a continuous domain (in our case, the time domain).
Like a multivariate
Gaussian, a GP is fully specified by a $(K \times 1)$ vector $\pmb{\mu}$ characterizing
the mean of the process and a $(K \times K)$
covariance matrix $\pmb{\Sigma}$ characterizing how the data points
are correlated with each other. To say that a measurement $\mathbf{f}$ is
``distributed as a GP'' means that we may write
%
\begin{align}
    \mathbf{f} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad,
\end{align}
%
i.e., the measurement is modeled as a random draw from a multivariate Gaussian.
GPs are therefore extremely easy to sample from.%
\footnote{Given a 1-d array \texttt{mean} and a 2-d array \texttt{cov} in \Python,
    sampling from the corresponding GP (if it exists)
    can be done in a single line of code by calling
    \texttt{numpy.random.multivariate\_normal(mean, cov)}.}
But the real showstopper is the application of GPs to inference problems.
Multivariate Gaussian distributions have a closed-form (marginal) likelihood
function, so it is easy (and extremely efficient) to compute the probability
of one's data conditioned on a given value of $\pmb{\mu}$ and $\pmb{\Sigma}$
(i.e., the ``likelihood''). This can in turn be maximized
to infer the optimal values of the model parameters
or used in a
numerical sampling scheme to compute the probability of those parameters
given the data (i.e., the ``posterior'').
Thanks to modern computer architectures, linear algebra packages, and
GP algorithms,
evaluating the GP likelihood may typically be done in a fraction of a second
for a reasonably-sized dataset (i.e., $K \lesssim 10^5$ datapoints).

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbf{f}$. Rather, GPs are a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$. A common approach to constructing
$\pmb{\Sigma}$ is to define a continuous \emph{kernel function} $k(t_i, t_j)$
such that the covariance between points at times $t = t_i$ and $t = t_j$ is
%
\begin{align}
    \Sigma_{i,j} = k(t_i, t_j)
    \quad.
\end{align}
%
In many applications, it is customary to further restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, which is typically chosen from a set of
standard functions with desirable smoothness and spectral properties. These
include, for example, the squared exponential kernel
%
\begin{align}
    k_\mathrm{E^2}(\Delta t) = \sigma^2 e^{-\frac{\Delta t}{2\tau}}
    \quad,
\end{align}
%
the Mat\'ern-3/2 kernel
%
\begin{align}
    k_\mathrm{M}(\Delta t) = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) e^{-\frac{\sqrt{3}\Delta t}{\tau}}
    \quad,
\end{align}
%
the exponential sine squared kernel,
%
\begin{align}
    k_\mathrm{S^2}(\Delta t) = \sigma^2 e^{-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)}
    \quad,
\end{align}
%
the cosine kernel,
%
\begin{align}
    k_\mathrm{C}(\Delta t) = \cos\left( \frac{2\pi \Delta t}{P} \right)
    \quad,
\end{align}
%
or any linear combination or product of those (as well as many other kernels not
listed here). In the expressions above, $\sigma$, $\tau$, $\Gamma$, and $P$ are
the \emph{hyperparameters} of the kernels, scalar quantities that (together with
the assumed functional form for the kernel) fully specify the covariance of the
Gaussian process.

>>>>

Over the past decade, Gaussian processes (GPs) have gained traction as
a leading tool for modeling correlated signals in astrophysical timeseries.
Despite whatever mystique the words ``Gaussian process'' may evoke, a GP
is nothing but a Gaussian distribution in many (formally infinite)
dimensions. Specifically, it is a Gaussian distribution over
\emph{functions} spanning a continuous domain (in our case, the time domain).
Like a multivariate
Gaussian, a GP is fully specified by a $(K \times 1)$ vector $\pmb{\mu}$ characterizing
the mean of the process and a $(K \times K)$
covariance matrix $\pmb{\Sigma}$ characterizing how the data points
are correlated with each other. To say that a measurement $\mathbf{f}$ is
``distributed as a GP'' means that we may write
%
\begin{align}
    \mathbf{f} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad,
\end{align}
%
i.e., the measurement is modeled as a random draw from a multivariate Gaussian.
GPs are therefore extremely easy to sample from.%
\footnote{Given a 1-d array \texttt{mean} and a 2-d array \texttt{cov} in \Python,
    sampling from the corresponding GP (if it exists)
    can be done in a single line of code by calling
    \texttt{numpy.random.multivariate\_normal(mean, cov)}.}
But the real showstopper is the application of GPs to inference problems.
Multivariate Gaussian distributions have a closed-form (marginal) likelihood
function, so it is easy (and extremely efficient) to compute the probability
of one's data conditioned on a given value of $\pmb{\mu}$ and $\pmb{\Sigma}$
(i.e., the ``likelihood''; see Equation~\ref{eq:log-like} below).
This can in turn be maximized
to infer the optimal values of the model parameters
or used in a
numerical sampling scheme to compute the probability of those parameters
given the data (i.e., the ``posterior'').
Thanks to modern computer architectures, linear algebra packages, and
GP algorithms,
evaluating the GP likelihood may typically be done in a fraction of a second
for a reasonably-sized dataset (i.e., $K \lesssim 10^5$ datapoints).

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbf{f}$. Rather, a GP
is a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$. Importantly, the covariance
$\pmb{\Sigma}$ of a GP is assembled via a continuous \emph{kernel function} $k(t_i, t_j)$
that specifies the covariance between points at times $t = t_i$ and $t = t_j$:
%
\begin{align}
    \Sigma_{i,j} = k(t_i, t_j)
    \quad.
\end{align}
%
In many applications, particularly when modeling stellar light curves,
it is customary to restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, typically chosen from a set of
standard functions with desirable smoothness and spectral properties.
Among a vast number of kernels are a few that are commonly used
in modeling stellar variability: the squared exponential kernel,
%
\begin{align}
    k_\mathrm{E^2}(\Delta t) = \sigma^2 e^{-\frac{\Delta t}{2\tau}}
    \quad,
\end{align}
%
the Mat\'ern-3/2 kernel,
%
\begin{align}
    k_\mathrm{M^\frac{3}{2}}(\Delta t) = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) e^{-\frac{\sqrt{3}\Delta t}{\tau}}
    \quad,
\end{align}
%
the exponential sine squared kernel,
%
\begin{align}
    k_\mathrm{S^2}(\Delta t) = \sigma^2 e^{-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)}
    \quad,
\end{align}
%
the cosine kernel,
%
\begin{align}
    k_\mathrm{C}(\Delta t) = \sigma^2 \cos\left( \frac{2\pi \Delta t}{P} \right)
    \quad,
\end{align}
%
or any linear combination and/or product of these. In the expressions above,
$\sigma$, $\tau$, $\Gamma$, and $P$ are
the \emph{hyperparameters} of the kernels.
Specifically, $\sigma^2$ is the variance (the value
of the kernel when $t_i = t_j$), $\tau$ is a typical timescale over which
data are correlated, $\Gamma$ is a dimensionless scale parameter, and $P$ is
a period. Once a kernel has been chosen, these scalar quantities fully
specify the covariance of the Gaussian process. GP-based inference
entails optimizing or sampling over these few quantities, which in many
cases can be done easily and efficiently.

Which brings us, finally, to an important drawback of GPs when applied
to modeling stellar variability. While it is straightforward to derive
posterior constraints on the hyperparameters of a GP, it is usually
not obvious what those constraints actually tell us about the star
or the actual source of the variability. While the GP period $P$
may usually be interpreted as the rotational period of the star,%
\footnote{An exception to this is in the presence of strong differential
    rotation, in which case many periods may be present in the data, or
    when spots evolve coherently, which can also introduce weak periodicities
    in the light curve.}%
the relationship between quantities like the amplitude $\sigma^2$
and the timescale $\tau$ and physical quantites of interest---such
as the properties of the star spots---are much fuzzier.
It may be tempting to interpret $\sigma^2$ as some measure of the
contrast or number of spots, or $\tau$ as the timescale on which the
spots evolve, but there are no guarantees these interpretations will
hold in general.
